The TFR-HDI controversy: an example Bayesian data analysis with PyMC

### 2011-07-10 tex/paper.tex
incremental improvements continue
nearing completion

\setlength{\topmargin}{0.0cm}
\setlength{\textheight}{21.5cm}
\setlength{\oddsidemargin}{0cm} 
\setlength{\textwidth}{16.5cm}
\setlength{\columnsep}{0.6cm}

@@

Implementing this in PyMC may seem hardly worth the trouble, since
this sort of ordinary least squares model has a simple closed-form
solution, as well as numerous implementations, like the regress
command in STATA.  It will be clear why this is worthwhile after
modifying the systematic part to yield a (slightly) more complicated
nonlinear model.

\begin{align*}
\text{Nonlinear Model (NM):}\qquad
\TFR_i &= \beta_0 + \beta_1 \HDI_i + \beta_2(\HDI_i - \gamma)^+ \ + \epsilon_i,\\
\epsilon_i &\sim N(0, \sigma^2),
\end{align*}
where $(\HDI_i - \gamma)^+$ equals $\HDI_i - \gamma$ if this quantity
is positive, and $0$ otherwise.

>>>>

\setlength{\topmargin}{0.0cm}
\setlength{\textheight}{21.5cm}
\setlength{\oddsidemargin}{0cm} 
\setlength{\textwidth}{16.5cm}
\setlength{\columnsep}{0.6cm}
\setlength{\parskip}{0.2cm}

@@

Implementing this in PyMC may seem hardly worth the trouble, since
this sort of ordinary least squares model has a simple closed-form
solution, as well as numerous implementations, like the regress
command in STATA.  It will be clear why this is worthwhile after
modifying the systematic part to yield a (slightly) more complicated
nonlinear model.  The code to implement it appears in
Fig.~\ref{src/linear model}.
\begin{figure}
\begin{verbatim}
import pylab as pl
import pymc as mc

import data

def linear():
    beta = mc.Uninformative('beta', value=[0., 0.])
    sigma = mc.Uniform('sigma', lower=0., upper=100., value=1.)

    @mc.deterministic
    def y_pred(beta=beta, X=data.hdi):
        return beta[0] + beta[1]*X
    y_obs = mc.Normal('y_obs', value=data.tfr,
                      mu=y_pred, tau=sigma**-2,
                      observed=True)

    return vars()

def fit_linear():
    vars = linear()

    mc.MAP(vars).fit(method='fmin_powell')

    m = mc.MCMC(vars)
    m.sample(iter=10000, burn=5000, thin=5)
    return m
\end{verbatim}
\caption{Listing of code to implement and fit linear model in PyMC}
\label{src/linear model}
\end{figure}

\begin{align*}
\text{Nonlinear Model (NM):}\qquad
\TFR_i &= \beta_0 + \beta_1 \HDI_i + \\
&\qquad \qquad\beta_2(\HDI_i - \gamma)^+ \ + \epsilon_i\\
\epsilon_i &\sim N(0, \sigma^2)
\end{align*}
where $(\HDI_i - \gamma)^+$ equals $\HDI_i - \gamma$ if this quantity
is positive, and $0$ otherwise.

Some exercises are to implement models for the following:
\begin{align*}
\text{Alt Linear Model:}\qquad
\log\TFR_i &= \beta_0 + \beta_1 \log(1-\HDI_i) + \epsilon_i\\
\epsilon_i &\sim N(0, \sigma^2)
\end{align*}

\begin{align*}
\text{Alt Linear Model 2:}\qquad
\TFR_i &= \beta_0 + \beta_1 \HDI_i + \beta_2 \HDI_i^2 + \epsilon_i\\
\epsilon_i &\sim N(0, \sigma^2)
\end{align*}

\begin{align*}
\text{Robust Linear Model:}\qquad
\TFR_i &= \beta_0 + \beta_1 \HDI_i + \epsilon_i\\
\epsilon_i &\sim L(0, \sigma^2)
\end{align*}
where $L$ is the Laplace distribution TK.
